---
title: Descriptive statistics by hand
author: Antoine Soetewey
date: '2020-01-08'
slug: descriptive-statistics-by-hand
categories: []
tags:
  - Statistics
meta_img: image/image.png
# description: Description for the page.
output:
  blogdown::html_page:
    toc: true
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = NA
)
```

# Introduction

This article explains how to compute the main descriptive statistics by hand. To learn how to compute these measures in R, read "[Descriptive statistics in R](/blog/descriptive-statistics-in-r/)".

Descriptive statistics (in the broad sense of the term) is a branch of statistics aiming at summarizing and describing a series of values or a dataset. Long series of values without any preparation or without any summary measures are often not informative due to the difficulty of recognizing any pattern in the data. Below an example with the height (in cm) of a population of 100 adults:

```{r, echo = FALSE, message = FALSE}
library(pander)
set.seed(42)
height <- round(rnorm(100, mean = 175, sd = 10), 2)
# range(height)
pander(height)
```

Facing this series, it is hard (not to say impossible) for anyone to understand the data and have a clear view of the size of these adults (in a reasonable amount of time). Descriptive statistics allow to summarize, and thus have a better overview of the data. Of course, by summarizing data through one or several measures, some information will inevitably be lost. However, in many cases it is generally better to lose some information but in return gain an overview.

# Location versus dispersion measures

Several different measures (called statistics) are used to summarize the data. Some of them give an understanding about the location of the data, others give and understanding about the dispersion of the data. In the following sections, we detail both location and dispersion measures and illustrate them with the data presented in the introduction (the height of the 100 adults).

## Location

Location measures allow to see "where" the data are located, around which values. In other words, location measures give an understanding on what is the central tendency or the "position" of the data. It includes the following statistics (others exist but we focus on the most common ones):

* minimum
* maximum
* mean
* median
* first quartile
* third quartile
* mode

We detail and compute (by hand) each of them in the following sections.

### Minimum and maximum

Minimum ($min$) and maximum ($max$) are simply the lowest and largest values, respectively. Given the height (in cm) of a sample of 6 adults:

```{r, echo = FALSE, message = FALSE}
sub_height <- head(height)
pander(sub_height)
```

The minimum is `r min(sub_height)` and the maximum is `r max(sub_height)`. These two simple measures give a clear idea about the heights of the smallest and tallest persons.

### Mean

The mean is probably the most common statistics. It gives an idea on what is the average value, that is, the central tendency of the data or the center of gravity. The mean is found by summing all values and dividing this sum by the number of observations (denoted $n$):

$$mean = \overline{X} = \frac{\text{sum of all values}}{\text{number of values}} = \frac{1}{n}\sum^{n}_{i = 1} X_i$$
Given our sample of 6 heights presented above, the mean is:

$$\overline{x} = \frac{188.7 + 169.3 + 178.6 + 181.3 + 179 + 173.9}{6}$$

(Note that a random variable is usually denoted with a capital letter and the particular values that our random variable can assume with a lowercase letter. That is the reason we used $\overline{X}$ and then $\overline{x}$.)

### Median

The median is another measure of location so it also gives an idea about the central tendency of the data. The easiest way to compute it is by first sorting the data from lowest to highest (in ascending order) then take the middle value. Formally, after sorting, the median is:

* if $n$ (number of observations) is odd: $$med(X) = X_{\frac{n+1}{2}}$$
* if $n$ is even: $$med(X) = \frac{1}{2}\big(X_{\frac{n}{2}} + X_{\frac{n}{2} + 1}\big)$$

where the subscript of $X$ denotes the numbering of the sorted data. The formulas look harder than they really are, so let's see with two concrete ewamples.

#### Odd number of observations

Given a sample of 7 heights taken from the 100 heights presented in the introduction:

```{r}
pander(tail(height, 7))
```

We first sort the order from lowest to highest:

```{r}
pander(sort(tail(height, 7)))
```

Given that the number of observations $n$ is odd (since $n = 7$), the median is

$$med(X) = X_\frac{7 + 1}{2} = X_4$$

So we take the fourth value from the sorted values, which corresponds to `r round(median(tail(height, 7)), 1)`. In conclusion, the median of `r pander(tail(height, 7))` is `r round(median(tail(height, 7)), 1)`.

#### Even number of observations

Now let's see when the number of observations is even, which is slighlty more complicated than when the number of observations is odd. Given a sample of 6 heights:

```{r}
pander(sub_height)
```

We sort the values in ascending order:

```{r}
pander(sort(sub_height))
```

Given that the number of observations $n$ is even (since $n = 6$), the median is

$$med(X) = \frac{1}{2}\big(X_{\frac{6}{2}} + X_{\frac{6}{2} + 1}\big) = \frac{1}{2}\big(X_{3} + X_{4}\big)$$

So we sum the third and fourth values from the sorted values and divide this sum by 2 (which is equivalent than taking the mean of these two middle values):

$$\frac{1}{2}(178.6 + 179) = 178.8$$

In conclusion, the median of `r pander(sub_height)` is `r round(median(sub_height), 1)`.

### Mean vs. median

The mean has the advantage that it gives a good picture of the data and it is usually understood by everyone. However, it has the disadvantage that the mean is sensible to outliers. To illustrate this, 

## Dispersion

* range
* standard deviation
* variance
* interquartile range

As for location measures, we now detail and compute (by hand) each of these statistics one by one.

### Range

The range is the maximum value minus the minimum value:

$$range = min - max$$


It is often the first step in many statistical project in the sense that it 

Thanks for reading. I hope this article helped you to understand the different descriptive statistics. If you would like to learn how to compute these measures in R, read "[Descriptive statistics in R](/blog/descriptive-statistics-in-r/)". As always, if you find a mistake/bug or if you have any questions do not hesitate to let me know in the comment section below, <a href="https://github.com/AntoineSoetewey/statsandr/issues" target="_blank" rel="noopener">raise an issue on GitHub</a> or [contact me](/contact/). Get updates everytime a new article is published by [subscribing to this blog](/subscribe/).
