---
title: Descriptive statistics by hand
author: Antoine Soetewey
date: '2020-01-02'
slug: descriptive-statistics-by-hand
categories: []
tags:
  - Statistics
meta_img: image/image.png
# description: Description for the page.
output:
  blogdown::html_page:
    toc: true
draft: true
---

# Introduction

This article explains how to compute the main descriptive statistics by hand. To learn how to compute these measures in R, read "[Descriptive statistics in R](/blog/descriptive-statistics-in-r/)".

Descriptive statistics (in the broad sense of the term) is a branch of statistics aiming at summarizing and describing a series of values or a dataset. Long series of values without any preparation or without any summary measures are often not informative due to the difficulty of recognizing any pattern in the data. Below an example with the height (in cm) of a sample of 100 adults:

```{r, echo = FALSE, message = FALSE}
library(pander)
set.seed(42)
height <- round(rnorm(100, mean = 175, sd = 10), 2)
# range(height)
pander(height)
```

Facing this sample, it is hard (not to say impossible) for anyone to understand the data and have a clear view of the size of these adults (in a reasonable amount of time). Descriptive statistics allow to summarize, and thus have a better overview of the data. Of course, by summarizing data through one or several measures, some information will inevitably be lost. However, in many cases it is generally better to lose some information but in return gain an overview.

# Location versus dispersion measures

Several different measures (called statistics) are used to summarize the data. Some of them give an understanding about the location of the data, others give and understanding about the dispersion of the data.

## Location

Location measures allow to see "where" the data are located, around which values. In other words, location measures give an understanding on what is the central tendency or the "position" of the data. It includes the following statistics:

* minimum
* maximum
* mean
* median
* first quartile
* third quartile
* mode

We detail and compute (by hand) each of them in the following sections.

### Minimum and maximum

Minimum (min) and maximum (max) are simply the lowest and largest values. Given the height (in cm) of a sample of 6 adults:

```{r, echo = FALSE, message = FALSE}
sub_height <- head(height)
pander(sub_height)
```

The minimum is `r min(sub_height)` and the maximum is `r max(sub_height)`. These two simple measures give a clear idea about the heights of the smallest and tallest persons.

### Mean

The mean is probably the most common statistics. It gives an idea on what is the average value that is used to derive the central tendency of the data. It is found by summing all values and dividing this sum by the number of observations ($n$):

$$mean = \bar{x} = \frac{\text{sum of all values}}{\text{number of values}} = \frac{1}{n}\sum^{n}_{i = 1} X_i$$
Given our sample of 6 heights presented above, the mean is:

$$\frac{188.7 + 169.3 + 178.6 + 181.3 + 179 + 173.9}{6}$$

## Dispersion

* range
* standard deviation
* variance

As for location measures, we now detail and compute (by hand) each of these statistics one by one.

### Range

The range is the maximum value minus the minimum value:

$$range = min - max$$


It is often the first step in many statistical project in the sense that it 

Thanks for reading. I hope this article helped you to understand the different descriptive statistics. If you would like to learn how to compute these measures in R, read "[Descriptive statistics in R](/blog/descriptive-statistics-in-r/)". As always, if you find a mistake/bug or if you have any questions do not hesitate to let me know in the comment section below, <a href="https://github.com/AntoineSoetewey/statsandr/issues" target="_blank" rel="noopener">raise an issue on GitHub</a> or [contact me](/contact/). Get updates everytime a new article is published by [subscribing to this blog](/subscribe/).
