---
title: How to embed a Shiny app in a Hugo website
author: Antoine Soetewey
date: '2020-01-07'
slug: how-to-embed-a-shiny-app-in-a-hugo-website
categories: []
tags:
  - Shiny
  - Web
meta_img: image/image.png
# description: Description for the page.
output:
  blogdown::html_page:
    toc: false
# draft: true
runtime: shiny
---

# Introduction

If you have developed a Shiny app and would like to embed it on your Hugo website, follow these steps:

1. create a new post as usual
1. add `runtime: shiny` in the YAML metadata
1. install the package "widgetframe" (`install.packages("widgetframe")`)
1. load the package in the body of your new post (`library(widgetframe)`)
1. insert the following code:

```{r, eval = FALSE, echo = TRUE}
# change URL with your URL
<iframe height="800" width="100%" frameborder="no" src="https://antoinesoetewey.shinyapps.io/optimal-asset-allocation/"> </iframe>
```

Here is an example with one of my Shiny app:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(reshape2)
library(shiny)
library(widgetframe)
library(knitr)
```

<iframe height="800" width="100%" frameborder="no" src="https://antoinesoetewey.shinyapps.io/optimal-asset-allocation/"> </iframe>